---
title: "Project Outline"
format: html
editor: visual
---

## Load Packages

```{r}
# Load necessary library 
library(tidyverse) 

library(dplyr) 

library(ggplot2)  
```

```{r}
# Global variable dataset is renamed to data for easier reference
data <- Sleep_Health_and_Lifestyle_Dataset
```

Using head(data) displays the first few rows of the dataset, providing a quick glimpse of the data structure and values.

Also, using summary(data)generates summary statistics for each variable in the dataset, including measures like mean, median, minimum, and maximum for numeric variables, and counts for factors or categorical variables.

```{r}
# View the first few rows of the dataset 
head(data)  

# Summary statistics of the dataset 
summary(data)


view(data)
# does this automatically change
```

In reviewing the initial dataset output from the `head(data)` command, I've decided to categorize the variables into two main types: categorical and continuous.

An outline of the general contents of each classification may be found below:

---
Categorical Variables:

1. Gender (Nominal): Indicates the gender, typically male or female.
2. Occupation (Nominal): Represents the job title or profession.
3. BMI.Category (Ordinal): Categories like underweight, normal, overweight, obese have an order based on BMI value.
4. Sleep.Disorde r(Nominal): Types of sleep disorders, such as None, Sleep Apnea,Insomnia, etc.
5. Blood.Pressure: Although it represents a measurement, it's often categorized based on ranges (normal, elevated, high), but the way it's formatted ('126/83') suggests it's being treated more as a continuous measure in practice. It might require preprocessing to be used effectively in either category.
---

---
Continuous Variables

1. Age (Continuous): Represents the age of individuals.
2. Sleep.Duration (Continuous): The amount of sleep in hours.
3. Quality.of.Sleep: This could be interpreted as either, but given it has a numeric scale, it's likely continuous or at least ordinal with a numeric scale.
4. Physical.Activity.Level (Continuous): Likely represents some quantifiable measure of physical activity.
5. Stress.Level: This appears to be on a numeric scale, so it could be continuous or ordinal.
6. Heart.Rate (Continuous): The number of heartbeats per minute.
7. Daily.Steps (Continuous): Number of steps taken in a day.
---

## **Visualizations**

By visualizing our dataset, we can provide insights into the relationships between variables, distribution of data points, and potential patterns or outliers.

Here are some visualization ideas and how you can implement them using ggplot2 in R:

### **1. Histograms for Continuous Variables**

-   Single Numeric Variable (Age Distribution): This code creates a histogram that shows the distribution of ages in the dataset. Each bar shows the number of people who fall into a given age range.

-   Single Numeric Variable (Sleep Duration Distribution): This code produces a histogram that shows the distribution of sleep durations. It offers insights into the sleep patterns within the dataset by showing the count of observations that fall within a given time range as bars.

```{r}
# Age Distribution
ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "seagreen", color = "black") +
  labs(title = "Distribution of Age", x = "Age", y = "Count")

# Sleep Duration Distribution
ggplot(data, aes(x = Sleep.Duration)) +
  geom_histogram(binwidth = 0.5, fill = "dodgerblue", color = "black") +
  labs(title = "Distribution of Sleep Duration", x = "Sleep Duration (hours)", y = "Count")

```

### **2. Boxplots for Continuous Variables by Categorical Groups**

-   This code allows for a comparison of the central tendency and variability of sleep quality across male and female participants by creating a boxplot that visualizes the distribution of sleep quality across different genders.

-   This code produces a boxplot that shows the distribution of heart rates across several BMI categories and offers information on the variability and central tendency of heart rates within each category.

```{r}
# Quality of Sleep by Gender
ggplot(data, aes(x = Gender, y = Quality.of.Sleep)) +
  geom_boxplot() +
  labs(title = "Quality of Sleep by Gender", x = "Gender", y = "Quality of Sleep")

# Heart Rate by BMI Category
ggplot(data, aes(x = BMI.Category, y = Heart.Rate)) +
  geom_boxplot() +
  labs(title = "Heart Rate by BMI Category", x = "BMI Category", y = "Heart Rate")

```

### **3. Scatter Plots for Relationships between Two Continuous Variables**

-   This code creates a scatter plot that shows the link between age and sleep quality, with each point denoting an individual's age and the associated score for quality of sleep. This allows for the investigation of any possible patterns or correlations between these variables.

-   This code generates a scatter plot that shows each person's daily steps and the related physical activity level score, enabling an analysis of the link between these variables. The scatter plot illustrates the relationship between the number of steps taken each day and the amount of physical activity.

```{r}
# Relationship between Age and Quality of Sleep
ggplot(data, aes(x = Age, y = Quality.of.Sleep)) +
  geom_point() +
  labs(title = "Age vs. Quality of Sleep", x = "Age", y = "Quality of Sleep")

# Daily Steps and Physical Activity Level
ggplot(data, aes(x = Daily.Steps, y = Physical.Activity.Level)) +
  geom_point() +
  labs(title = "Daily Steps vs. Physical Activity Level", x = "Daily Steps", y = "Physical Activity Level")

```

### **4. Bar Plots for Categorical Variables**

-   This code produces a bar plot that counts the number of people in each BMI group, giving the distribution of people across various BMI categories a visual depiction.

-   This code generates a bar plot, where each bar represents the mean stress level for a particular occupation, to show the average stress level across various occupations. Whereas the y-axis shows the average stress level, the x-axis indicates the occupation. To improve legibility, the x-axis labels have also been angled in the theme.

```{r}
# Count of Individuals by BMI Category
ggplot(data, aes(x = BMI.Category)) +
  geom_bar(fill = "mediumorchid") +
  labs(title = "Count of Individuals by BMI Category", x = "BMI Category", y = "Count")

# Average Stress Level by Occupation
ggplot(data, aes(x = Occupation, y = Stress.Level)) +
  geom_bar(stat = "summary", fun = "mean", fill = "orange") +
  labs(title = "Average Stress Level by Occupation", x = "Occupation", y = "Average Stress Level") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### **5. Faceted Plots for Multi-variable Analysis**

-   This code creates a scatter plot, where each point represents a person's daily steps and associated physical activity level score, illustrating the relationship between daily steps and physical activity level. By breaking down the plot into BMI categories, it becomes possible to compare these connections between various BMI ranges.

```{r}
# Scatter Plot of Daily Steps vs. Physical Activity Level, Faceted by BMI Category
ggplot(data, aes(x = Daily.Steps, y = Physical.Activity.Level)) +
  geom_point() +
  facet_wrap(~ BMI.Category) +
  labs(title = "Daily Steps vs. Physical Activity Level by BMI Category", x = "Daily Steps", y = "Physical Activity Level")

```

## Explore The Dataset

-   The `glimpse()` method shows the variable names and data types associated with the first few rows, giving a brief summary of the dataset. This makes it possible to quickly comprehend the contents and structure of the dataset.

```{r}
glimpse(data)
```

-   This code calculates and prints the mean age of individuals in the dataset.

-   This code calculates and prints the mean sleep duration of individuals in the dataset.

-   This code calculates and prints the mean sleep quality of individuals in the dataset.

```{r}
print(paste("Mean Age:", round(mean(data$Age))))
print(paste("Mean Sleep Duration:", round(mean(data$Sleep.Duration))))
print(paste("Mean Sleep Quality", round(mean(data$Quality.of.Sleep))))
```

-   This code provides summary statistics, such as mean, median, minimum, maximum, and quartiles, for the numerical variables in the dataset.

```{r}
# Summary statistics for numerical variables
summary(data)
```

-   These lines produce frequency tables that display the counts of each category inside the categorical variables "Gender," "Occupation," "BMI.Category," and "Sleep.Disorder."

```{r}
# Counts for categorical variables
table(data$Gender)
table(data$Occupation)
table(data$BMI.Category)
table(data$Sleep.Disorder)
```

### **Histograms for Numerical Variables:**

```{r}
#Visualize the distributions of numerical variables using histograms.

# Histogram for Sleep Duration
hist(data$Sleep.Duration, main = "Sleep Duration Distribution", xlab = "Sleep Duration (hours)")

# Histogram for Quality of Sleep
hist(data$Quality.of.Sleep, main = "Quality of Sleep Distribution", xlab = "Quality of Sleep")

```

### **Boxplots for Categorical Variables:**

```{r}
#Compare distributions of numerical variables across different categories of categorical variables using boxplots.

# Boxplot of Sleep Duration by Gender
boxplot(Sleep.Duration ~ Gender, data = data, main = "Sleep Duration by Gender", xlab = "Gender", ylab = "Sleep Duration")

# Boxplot of Quality of Sleep by BMI Category
boxplot(Quality.of.Sleep ~ BMI.Category, data = data, main = "Quality of Sleep by BMI Category", xlab = "BMI Category", ylab = "Quality of Sleep")
```

### **Visualize Correlations:**

```{r}
# Scatter plot for Age and Sleep Duration
ggplot(data, aes(x = Age, y = Sleep.Duration)) +
  geom_point() +
  labs(title = "Scatter Plot of Age vs Sleep Duration", x = "Age", y = "Sleep Duration")
```

```{r}
# Scatter plots for Age vs. Quality of Sleep, Age vs. Physical Activity Level, and Quality of Sleep vs. Physical Activity Level
ggplot(data, aes(x = Quality.of.Sleep, y = Physical.Activity.Level, color = Age)) +
  geom_point() +
  labs(title = "Scatter Plot of Quality of Sleep vs Physical Activity Level", x = "Quality of Sleep", y = "Physical Activity Level")

```

```{r}
ggplot(data, aes(x = Age, y = Physical.Activity.Level, color = Quality.of.Sleep)) +
  geom_point() +
  labs(title = "Scatter Plot of Age vs Physical Activity Level", x = "Age", y = "Physical Activity Level")
```

```{r}
ggplot(data, aes(x = Age, y = Quality.of.Sleep, color = Physical.Activity.Level)) +
  geom_point() +
  labs(title = "Scatter Plot of Age vs Quality of Sleep", x = "Age", y = "Quality of Sleep")
```

## Cleanse The Dataset

```{r}
#Number of missing values in each attribute
colSums(is.na(data))
```

```{r}
#Use boxplots to check for outliers within numeric values
data |>
  ggplot()+
  geom_boxplot(aes(y = Age))

data |>
  ggplot()+
  geom_boxplot(aes(y = Sleep.Duration))

data |>
  ggplot()+
  geom_boxplot(aes(y = Quality.of.Sleep))

data |>
  ggplot()+
  geom_boxplot(aes(y = Physical.Activity.Level))

data |>
  ggplot()+
  geom_boxplot(aes(y = Stress.Level))

data |>
  ggplot()+
  geom_boxplot(aes(y = Heart.Rate))# Heart rate does have outliers. remove?

data |>
  ggplot()+
  geom_boxplot(aes(y = Daily.Steps))
```

Investigating the Heart Rate Outliers

```{r}
# How many people have different outlier heart rates
data[data$Heart.Rate > 79,] |>
  ggplot() +
  geom_bar(aes(x = Heart.Rate))

# Exact number of outlier heart rates
heartRateOutlierCount <- nrow(data[data$Heart.Rate > 79,])
print(heartRateOutlierCount)
# 15 out of 374 entries

# Looking at the instances of heart rates over 79
print(data[data$Heart.Rate > 79,])

# It seems like the outliers have a commonality: high blood pressure
# High blood pressure can cause increase in heart rate, there can also be other factors
ggplot(data, aes(x = Heart.Rate, y = Blood.Pressure)) +
  geom_point()

```

## Initial Questions

1\) Which occupation experiences the best/worst sleep quality?

2\) Do people with higher levels of activity have lower stress levels than people with lower activity levels?

3\) What is the distribution of sleep duration among different age groups and genders?

4\) How does the quality of sleep vary with age, occupation, and physical activity level?

5\) Are there any correlations between physical activity level, stress level, and sleep quality?

6\) Do individuals with certain occupations or lifestyles have higher rates of sleep disorders?

7\) What are the common factors associated with high or low quality of sleep?

8\) How do sleep duration and quality of sleep relate to other health metrics such as BMI, blood pressure, and heart rate?

## Analyze Variance

```{r}
#bar chart to visualize variance between male and female values
data|>
  ggplot()+
  geom_bar(aes(x = Gender))

#bar chart for Occupation values
data|>
  ggplot()+
  geom_bar(aes(y = Occupation))

data|>
  ggplot()+
  geom_bar(aes(x = BMI.Category))

#Use bar chart for Blood pressure
data|>
  ggplot()+
  geom_bar(aes(y = Blood.Pressure))

#Use bar chart to check for outliers in Blood pressure
data|>
  ggplot()+
  geom_bar(aes(y = Sleep.Disorder))
```

## Calculate Covariance Among Attributes

```{r}
covariance_matrix <- cov(data[sapply(data, is.numeric)])
print(covariance_matrix)

```
